### 2023-11-21 response_to_r1.md

# Response to reviewers of "Spreadsheet modeling and wrangling with Python"

I'd like to thank the reviewer and the editor for their thoughtful comments on this paper. I have
done by best to address them.

## Editor comments

**Comment 1**

I would like to apologize to the author for the delay in the review.  Although two reviewers accepted the assignment, one of them, at this point, is 65 days late with the review – therefore, I decided to proceed with a single review.

My first request to the author is to include other related ITE publications either as part of the literature review or to the other relevant sections.  Here is a short list of links to start:
•       https://pubsonline.informs.org/doi/pdf/10.1287/ited.2021.0261
•       https://pubsonline.informs.org/doi/full/10.1287/ited.2023.0285
•       https://pubsonline.informs.org/doi/pdf/10.1287/ited.2022.0275

**Response 1**

In the opening paragraph I refer to the seminal issues of ITE but didn't reference specific papers. I've rectified that and have added the following references.

- Alderson DL (2022) Interactive computing for accelerated learning in computation and data science. INFORMS Transactions on Education 22(2):130–145, URL http://dx.doi.org/10.1287/ited.
2021.0261.
- Baker K (2000) Gaining insight in linear programming from patterns in optimal solutions. INFORMS Transactions on Education 1(1):4–17, URL http://dx.doi.org/10.1287/ited.1.1.4.
- Bell PC (2000) Teaching business statistics with microsoft excel. INFORMS Transactions on Education 1(1):18–26, URL http://dx.doi.org/10.1287/ited.1.1.18.
- Carraway RL, Clyman DR (2000) Integrating spreadsheets into a case-based mba quantitative methods course: Real managers make real decisions. INFORMS Transactions on Education 1(1):38–46, URL http://dx.doi.org/10.1287/ited.1.1.38.
- Evans JR (2000) Spreadsheets as a tool for teaching simulation. INFORMS Transactions on Education 1(1):27–37, URL http://dx.doi.org/10.1287/ited.1.1.27.
- Isken MW (2003) Data cleansing and analysis as a prelude to model based decision support. INFORMS Transactions on Education 3(3):23–75, URL http://dx.doi.org/10.1287/ited.3.3.23.
- Isken MW (2014) Translating a lab based spreadsheet modeling course to an online format: Experience from a natural experiment. INFORMS Transactions on Education 14(3):120–128, URL http://dx.doi.org/10.1287/ited.2013.0123.
- Powell SG (2001) Teaching Modeling in Management Science. INFORMS Transactions on Education 1(2):62–67, ISSN 1532-0545, URL http://dx.doi.org/10.1287/ited.1.2.62.
- Ragsdale CT (2001) Teaching management science with spreadsheets: From decision models to decision support. INFORMS Transactions on Education 1(2):68–74, URL http://dx.doi.org/10.1287/ited.1.2.68.
- Savage S (2001) Blitzograms—interactive histograms. INFORMS Transactions on Education 1(2):77–87, URL http://dx.doi.org/10.1287/ited.1.2.77.

The publication by Alderson (2022) is very relevant and I'm not sure how I missed seeing that. I've integrated it into the paper and cited it. Thank you for bringing it to my attention.

"One particularly relevant example are the courses in computation offered by the Naval Postgraduate School and described in Alderson (2022). In their courses, Jupyter notebooks (discussed in more detail in the next section) play a prominent role in providing an interactive development and learning environment for Python within an operations research context."

In the first version I avoid citing my own work in ITE even though it is quite relevant. It now seems appropriate now to add these and I have done
that in the section describing my speadsheet based analytics course.

**Comment 2**

Second, I found it difficult to follow the structure of the environment within which the course it taught.  Could the author please include a graphic or a table to add a visualization to the explanation of where in the program’s curriculum the AAP course fits in, and where in the AAP course the EwP module fits in.

**Response 2**

I understand. I've added more explanation to the relevant paragraph. As opposed to adding a table or graphic that abstracts the details, I'm encouraging the readers to visit the relevant course landing pages as they include nicely structured topic outlines that make it very clear how EwP fits in to the Python material covered by our PCDA and AAP courses.

Here's the reworked paragraph:

"At our institution, The EwP module is part of a business course entitled  [Advanced Analytics with
Python](http://www.sba.oakland.edu/faculty/isken/courses/aap) (AAP). The students taking this course have already taken the PCDA course within which they learned fundamental Python programming within the context of data analytics over seven weeks. The PCDA course ends with intro to the [scikit-learn library](https://scikit-learn.org/stable/index.html) for doing predictive modeling in Python and the AAP course begins where the PCDA course ends. The EwP module follows immediately after the modules on machine learning and builds on the Python fundamentals which were covered in the PCDA course. The EwP module could also be used as part of a semester long Python based analytics course in which the Python modules covered in the PCDA course are followed by more advanced topics such as EwP or other topics in our AAP course. The only reason the EwP module is not included in the PCDA course is that Python is only covered in half of that course - the other half being an introduction to R and Linux. If you visit the open access [PCDA course
website](http://www.sba.oakland.edu/faculty/isken/courses/pcda/)  and [AAP course
website](http://www.sba.oakland.edu/faculty/isken/courses/aap/), you can see the topic outline for each course and how the EwP module is positioned."

I'd like to stress that a big part of my paper is that it isn't just a paper. It is backed by a huge amount of public, freely available and Creative Commons licensed course websites and course content. Anyone who is interested in adapting my materials for their course is free to do so. Everything other than all of the homework assignments are publicly accessible from the course websites. That's why throughout this paper I provide hyperlinks to relevant parts of the EwP course website or course content. This allows the reader to have a richer reading experience - instead of just a paper which necessarily is limited by what it can contain, the reader can view the entirety of the Jupyter notebooks, in html format, which make up the modules. ITE is an online journal and I'm trying to take advantage of the capabilities offered by such a platform.

**Comment 3**

Are the students already familiar with the Jupyter notebook environment from their PCDA course, and therefore the instructor continues using the same environment, or are the students using a different development environment in their preceding course?

**Response 3**

Yes, we use Jupyter notebooks as well as a integrated development environment (either Spyder, PyCharm, or VSCode) in the PCDA course. The same tools are used in the AAP course. All of the technical and logistical details of both courses are described in excruciating detail in the public course websites linked to in the paper.

**Comment 4**

The same with source control – it is not clear to me why the program chose to include software engineering related discussion about source control in the EwP module of the AAP course, could the author please add a bit of discussion around that?

**Response 4**

I didn't choose that. Version control is covered in the very first module of the AAP course and students use git and GitHub for the ML module that precedes the EwP module. The following sentence is in the original version of the paper:

"Students are reminded of the importance of version control and we initialize a git repository for our budding package. The basics of version control were already covered in the first week of the course."

Again, the AAP course website main page at [http://www.sba.oakland.edu/faculty/isken/courses/aap/](http://www.sba.oakland.edu/faculty/isken/courses/aap/) contains a detailed topic outline and the reader can see what is covered when.

**Comment 5**

Third, would it be possible to add the learning goals of the EwP module and suggested methods for assessments?

**Response 5**

Each submodule of EwP has a number of learning objectives that appear at the top of each
submodule course webpage. I've added screenshots of these objectives so that readers can see
how the objectives are presented to the students.

As for assessment, in the final section of the paper, *Classroom experience and the future*, we described and made available for download an assignment that we have used for assessment of this material. It covers all aspects of the EwP module.

**Comment 6**

Fourth, the author should also address the comments and suggestions provided by Reviewer 1.

Thank you for the submission to ITE and sharing your ideas and experiences with our academic community.  My apology again for the delay in the review process.

**Response 6**

See below for comments and responses to Reviewer 1.

## Reviewer #1 comments

Thank you taking the time to review my paper and to provide thoughtful feedback.

**Comment 1**

In the Introduction section, there appears to be too much emphasis on offering
the EwP approach as an alternative to Excel due to its limitations. I don’t think
many folks will buy the argument. The discussion seems to shortchange Excel’s
ability to set up connections with externally managed data while maintaining the
flexibility and ease of summarizing data in Excel. An alternative is to simply
suggest previous work in Excel by students provides a good launch point for
developing the more advanced programming skills in Python.

**Response 1**

I'm definitely not trying to offer EwP as an alternative to Excel. I've taught an Excel based spreadsheet modeling course for over 20 years and have used Excel extensively in both industry as well as in academic settings. I love spreadsheets and have for a long time. But, they do
have their limitations and the better our students are at assessing when a spreadsheet is and is not appropriate, the better off they will be. This is not some novel argument. Excel bashing is rampant in corporate IT and analytics settings. Quite honestly, even Jupyter notebooks have major shortcomings when it comes to productionizing analytical models implemented in Python and I mention this in the paper - and that's why I also make students use an actual IDE to write code in addition to the things we do in notebooks. There's a lot of nuance in knowing how to effectively wield tools like Excel and Python in professional analytics environments. This is a big part of the motivation behind the development of my courses and in writing this paper.

My intent is to provide an interesting way for business students to learn more advanced Python concepts through a very familiar lens of spreadsheet modeling. In addition, I want to expose them to the abilities of Python in automating manipulation of spreadsheets. Now with MS announcing the inclusion of Python within Excel, the complementary nature of these two powerhouse tools will become even more important. To that end, I've modified the beginning of that paragraph as follows:

"Despite numerous calls for, and predictions of, the eradication of spreadsheets from the business world, Excel and other spreadsheets are \href{https://benn.substack.com/p/the-next-billion-programmers}{alive and well}. Their flexibility is unparalleled and it is unrealistic to expect them to disappear anytime soon. If anything, the [recent announcement by Microsoft](https://techcommunity.microsoft.com/t5/excel-blog/announcing-python-in-excel-combining-the-power-of-python-and-the/ba-p/3893439)
 that users will be able combine Python scripts with Excel formulas within the same workbook, will likely solidify Excel's place in the analytics space."

**Comment 2**

In sections 3.2 and 3.3, it would have been nice to see more figures illustrating
the coding associated with Goal Seek and Monte Carlo simulation exercises.

**Response 2**

I wrestled with this when writing the paper. Instead of including more figures, I opted, as I described above, to include prominent hyperlinks to HTML versions of the notebooks in the relevant section headers.

For example:

3.3. Notebook 3: Monte-Carlo simulation (html version)

I also added explicit language early in the paper that highlighted these hyperlinks and encouraged readers to use them.

"While all of the notebooks are publicly available (see Section 6), links to html versions of
each of the notebooks are provided throughout the document and readers are encouraged to use
them while reading to see exactly how each topic is presented."

My intent is for the reader to open this html version in a browser tab and then they can see the entire notebook. Code snippets taken out of context are not nearly as useful as following the narrative flow of the notebook - this is the real strength of Jupyter notebooks for teaching.

If the reviewers really want me to add more code screenshots, I can, but I'd really prefer readers look at the (html versions of) the actual notebooks. Of course, the actual Jupyter notebooks are also available from the course website.

**Comment 3**

In section 7, the author(s) need to more fully address the roll out of Python in
Excel. Does this alter their pedagogy? It may not if the focus is develop the skills
outlined in the Introduction and listed below
o  teaching relevant and more advanced Python to business students using familiar and
simple modeling examples,  conveying a sense of the software design and development
process,  acting as a tour guide for a learning journey,  teaching students to think and
act like a software developer, not just an analyst
In any case, I think the paper needs a more comprehensive discussion regarding
this development.

**Response 3**

I've added reference to this development in the opening section as I mentioned in Response 1 above. I've also added a bit in the last section:

"Rumors have swirled for a number of years that Microsoft was considering adding Python as a first class language alternative to VBA within their MS Office suite of packages \cite{cimpanuMicrosoftConsidersAdding2017}. During the review process for this article, this development [finally
materialized](https://techcommunity.microsoft.com/t5/excel-blog/announcing-python-in-excel-combining-the-power-of-python-and-the/ba-p/3893439)
, albeit in a limited roll out to beta testers. This has potentially enormous implications for the future of both Excel and Python. At this point, it is probably safe to say that this development further underscores the importance of business analytics students adding Python to their set of analytical toolbox."

Other than that, I'm not in a position to say a whole lot more right now. The roll out was restricted to those with Beta Channel access (which I do not have). It appears right now that this feature is designed to run in the Azure cloud environment, which is a little concerning. It's also not clear if we'll be able to write full blown code modules in Python instead of Excel VBA. It's just too early to know how this is going to play out.
If and when this development is widely released, then certainly I will develop a teaching module to add to my AAP class and will likely add a very introductory module to my spreadsheet based analytics course. This will be a rich area for collaboration between the Excel and Python communities. Pretty exciting development but pales a bit in comparison to the chaos around Microsoft and OpenAI right now. :)

**Editing Comments**

On page 5, 2 nd paragraph – The sentence “ After running the cells, they appear as show
in Figure 2.” – The word show should be changed to shown

**Response**

Oops. It's fixed now.

### 2023-11-21 python_excel_ite_r1.tex
finished response to reviewers

Despite numerous calls for, and predictions of, the eradication of spreadsheets from the business world, Excel and other spreadsheets are \href{https://benn.substack.com/p/the-next-billion-programmers}{alive and well}. Yes, spreadsheets have well known limitations. Spreadsheets are not databases and probably should not be used as one, especially for relational data models. There are better data visualization and exploration tools. There are better dashboarding tools that are more easily maintained in a production environment. Spreadsheet errors can be a \href{https://eusprig.org/research-info/horror-stories/}{huge problem}. Version control is difficult as both data and logic live in the same spreadsheet document. Documentation of multi-step manual data manipulation processes is often lacking -- hindering reproducibility and automation. There are data size limitations and external data links present their \href{https://www.thespreadsheetguru.com/find-remove-external-links/}{own set of challenges}. But, Excel continues to be used for a wide range of quantitative analysis related activities throughout the business world. One particular use is the building and exercising of models of business problems
to explore the impact of important model inputs on different output metrics. This ``what if?'' type modeling is facilitated by Excel's flexibility, numerous built in functions, integrated sensitivity analysis via \href{https://support.microsoft.com/en-us/office/calculate-multiple-results-by-using-a-data-table-e95e2487-6ca6-4413-ad12-77542a5ea50b}{Data Tables},  \href{https://support.microsoft.com/en-us/office/use-goal-seek-to-find-the-result-you-want-by-adjusting-an-input-value-320cb99e-f4a4-417f-b1c3-4f369d6e66c7}{Goal
	Seek} for break-even analysis and exploratory data analysis via plots and Pivot Tables.

@@

Most AAP students have also already taken   \href{http://www.sba.oakland.edu/faculty/isken/courses/ba}{Business Analytics} (BA) which is a spreadsheet based introduction to business analytics (\cite{isken2003:muddy,isken2014:lab}). The BA and PCDA courses have historically been offered in face to face mode in a computer teaching lab. Since 2020, these two courses have been also offered as online, asynchronous courses. The newer AAP course has been offered since 2021 in an online asynchronous mode. All three courses have extensive course websites that are publicly accessible, including access to all of the video content and supporting files. These courses serve upper level undergraduate and graduate students. Many are from the School of Business Administration but they draw students from many programs across our campus.

@@

All of the Python modules in the PCDA course as well as the entire AAP course are taught using a combination of Jupyter notebooks, Python scripts, and videos that walk readers through the notebooks. Jupyter notebooks, part of \href{https://jupyter.org/}{Project Jupyter}, are an example of \textit{literate computing} (\cite{perez2015project}) and allow one to create rich interactive documents that support iterative exploration and development.

Central to Jupyter notebooks is a mix of \textit{code cells}, \textit{markdown cells} and \textit{output cells}. Code cells contain executable code while markdown cells contain plain text using \href{URL}{markdown} for styling text. For example, Figure \ref{fig:cells_edit_mode} shows two markdown cells separated by a code cell as they look when in editing mode. After running the cells, they appear as show in Figure \ref{fig:cells_after_running}. After executing a code cell, the output appears directly below it in an output cell. Mixing code and markdown cells makes it quite easy to create interactive tutorials in which students can not only run pre-written code, but can also add their own code -- all with the aid of supporting explanatory text. All of the blog posts upon which EwP is based were written in Jupyter notebooks.

The ability to weave standard explanatory language with formal programming languages has made notebook computing a popular tool in the data science education community (\cite{perkel2018jupyter}). Jupyter notebooks can also include equations written in \LaTeX, data visualizations, and other multi-media. The technology is free, open source, runs on most computing platforms, and has a huge user community that has created a large number of resources for students, faculty and industry practitioners. There are countless introductory tutorials for Jupyter notebooks, but the official Project Jupyter Documentation (\cite{jupyterteamProjectJupyterDocumentation2015}) is a very good place to start if you are new to notebook computing. As with all tools, there are pros and cons to their use. See \cite{rule2018ten, barba2019teaching} and \cite{johnson2020benefits} for guidance on Jupyter notebook best practices for both research and teaching.

@@

There are two overarching learning objectives for this \href{http://www.sba.oakland.edu/faculty/isken/courses/aap/mod3a_whatif.html}{first EwP submodule} - see Figure \ref{fig:objectives_whatif} for how the objectives are presented to students via the course website. Students learn how Python might be used to do something they are quite familiar with doing in a spreadsheet -- model building and sensitivity analysis. At the same time, students encounter, as needed, more advanced Python programming concepts and techniques. In this way, the computational problem to be solved drives the introduction of more advanced Python concepts and techniques instead of such things being presented in a vacuum. I make it a point to show that there are often multiple possible approaches to a computational problem by being willing to abandon one approach and trying another. Emphasis is given to the lessons learned from each path explored. Every notebook is accompanied by screencasts that walk the students through the material and act as a guided tour with challenges to overcome along the way. While all of the notebooks are publicly available (see Section \ref{sec:software}), links to an html version of each of the notebooks are provided throughout the document.

@@

The stage is set with an overview of a typical spreadsheet based model that most of the students have seen before. The notebook also includes a screenshot (Figure \ref{fig:xl_model}) of this Excel model. Readers are encouraged to click the \href{http://www.sba.oakland.edu/faculty/isken/excel_with_python/what_if_1_model_datatable.html}{html version} link so that you can see all of the notebook details as you are reading through this section.

@@

We then introduce a similar modeling problem that will be the focus of the rest of the notebook. The Walton Bookstore problem appears in the chapter on Monte-Carlo simulation in textbooks that I have used in my spreadsheet modeling course \cite{albrightBusinessAnalyticsData2016,winstonPracticalManagementScience2018}. The (slightly modified) problem is as follows:

@@

At this point we discuss a fundamental difference between the computing model of Excel and that of Jupyter notebooks. While spreadsheets respond automatically to changes, Jupyter notebooks require rerunning all code cells that include or follow the changed code. This difference in behavior between Jupyter notebooks and spreadsheets might be one of the most important concepts for students to internalize. A well known challenge, often referred to as the \textit{hidden state problem}, associated with computational notebooks is the ease of running code cells out of order and introducing inadvertent errors -- see \cite{johnson2020benefits,grusDonNotebooks2018}. This point along with strategies to avoid it are discussed during the first week of the course which focuses on \href{http://www.sba.oakland.edu/faculty/isken/courses/aap/jupyter_notebooks.html}{notebook computing best practices as well as a historical look} at computational notebooks.

@@

Since order quantity is the key decision variable, we might want to see how profit changes for different order quantities. In Excel, the Data Table tool provides a familiar and easy way to do this. In a Data Table, a range of order quantities can be used as row or column input and one or more output values can be computed. In Python, the vectorized nature of the \href{https://numpy.org/}{NumPy} library (\cite{harris2020array}) provides a simple way to accomplish the same thing. A range of order quantities is created as a NumPy array that can be used directly to compute vectors of all of the intermediate and final output variables. Much like Data Tables, vectorized computations avoid having to explicitly iterate, or loop, through a collection of input values.

@@

Before trying to build our \code{goal\_seek} function we take a brief detour into basic root finding algorithms through packages like \href{https://docs.scipy.org/doc/scipy/tutorial/optimize.html#root-finding}{SciPy} (\cite{2020SciPy-NMeth}) as well as in \href{https://github.com/patrickwalls/mathematical-python/}{open source notebooks and scripts} that are widely available in the Python data science ecosystem (\cite{wallsMathematicalPython2023}). We compare the output of Excel's Goal Seek with different starting values to the values obtained by different root finding algorithms implemented in Python. The importance of starting values and also the range of different root finding algorithms available is often surprising to business students who have spent most of their analytical life in Excel.

@@

In the Excel modeling world, Monte-Carlo simulation can be done without add-ins, but packages like \href{https://lumivero.com/products/at-risk/}{@Risk} can make the process much easier. In particular, random variable generation for a wide range of probability distributions is facilitated by add-ins like @Risk. In Python, the \href{https://numpy.org/doc/stable/reference/random/generator.html}{NumPy package provide random variate generation} for a large number of distributions and similarly, \href{https://docs.scipy.org/doc/scipy/reference/stats.html}{SciPy provides functions for computing distribution related quantities} for many probability distributions. Much like we did when creating the \code{data\_table} function, we start with adding uncertainty to a single input variable in the bookstore model and rely on NumPy's vectorized computing capabilities to generate the profit associated with each realization of the random input. Then we can do standard statistical analysis of the simulation output using Python packages such as pandas and SciPy. In spreadsheet based simulation textbooks, the notion of the ``Flaw of Averages'' popularized by Sam Savage (\cite{savageFlawAveragesWhy2012}) is often illustrated. We do the same with our Python based model, showing how replacing random demand by its mean can result in a wildly optimistic estimate of the mean profit found through analyzing the simulation output. Much like doing simulation in Excel without add-ins, using Python makes transparent the quite simple and brute force nature of Monte-Carlo simulation.

@@

Now that we have created some useful modeling functionality, our goal is to make it easier for a modeler to actually use the code. Certainly we do not want to rely on copying, pasting and modifying code when building a new model. In the Python world, the natural next step is to deploy our code as a Python \textit{package} (\cite{buezenWelcomePythonPackages2022}) that can then be imported like any other Python library. Figure \ref{fig:objectives_packaging} shows the learning objectives from this \href{http://www.sba.oakland.edu/faculty/isken/courses/aap/mod3b_whatif_packaging.html}{submodule page on the course website}.

@@

Creating a deployable Python package is a good exercise in designing and creating a reusable software artifact. Thinking about how others will interact with your software usually leads to better designed software.  In this submodule we create a Python package that can be imported and used by other modelers. This will require some design changes to our software and will also involve moving code out of Jupyter notebooks and into Python script files. While Jupyter notebooks can be integrated into production environments \cite{uffordInteractiveNotebookInnovation2019}, it is much more common for code to live in Python script files. In addition to creating a deployable package, this submodule also discusses the importance of documentation in its many forms in a Python project.

@@

The first notebook in this submodule starts by explaining what Python packages are and their role in sharing code. We then revisit a concept from earlier in the course -- creating a good project folder structure. For this we use what is known as a \href{https://cookiecutter.readthedocs.io/en/stable/}{cookiecutter} that can automatically generate a project folder structure and key files from the answers to a few prompts (\cite{audreyroyandcookiecuttercommunityCookiecutterBetterProject2012}). There are a few different folder layouts used in Python projects and there is a bit of discussion on the folder structure we will use and resources given for exploring this issue further. Students are reminded of the importance of version control and we initialize a git repository for our budding package. The basics of version control were already covered in the first week of the course. My experience is that most students have little exposure to software project management concepts such as project folder structures and version control. This can lead to a jumble of files with incoherent names used as a proxy for version control -- \code{whatif\_v1.py}, \code{whatif\_v2.py}, \code{whatif\_final.py}, \code{whatif\_finalfinal.py}, and so on.

@@

This notebook ends by pointing the students to another notebook in which our newly deployed whatif package is used for a completely different model -- the New Car Simulation that has been a part of Winston and Albright's textbooks for many years (\cite{winstonPracticalManagementScience2018,albrightBusinessAnalyticsData2016}). This new model differs structurally from the Bookstore Model, but is representative of many financial models in that cash flows over multiple years need to be computed and summarized. This helps cement the idea that there is value in creating reusable code.

@@

This deployment focused submodule ends with a short notebook discussing the different types of documentation needed in a typical project. This includes code comments, \href{https://peps.python.org/pep-0257/}{docstrings}, \code{readme} files and generating documentation by creating \href{https://www.sphinx-doc.org/en/master/usage/restructuredtext/basics.html}{reStructuredText} files (similar to but more powerful than markdown) and using \href{https://www.sphinx-doc.org/en/master/index.html}{Sphinx}, the Python documentation generation tool. This is one of those tasks that few people like to do but can be very important for the long term success of a project. Even if you are just creating a tool for yourself, you will forget the details over time and documentation can be very beneficial in refreshing your memory.

>>>>

Despite numerous calls for, and predictions of, the eradication of spreadsheets from the business world, Excel and other spreadsheets are \href{https://benn.substack.com/p/the-next-billion-programmers}{alive and well}. Their flexibility is unparalleled and it is unrealistic to expect them to disappear anytime soon. If anything, the \href{https://techcommunity.microsoft.com/t5/excel-blog/announcing-python-in-excel-combining-the-power-of-python-and-the/ba-p/3893439}{recent announcement by Microsoft} that users will be able combine Python scripts with Excel formulas within the same workbook, will likely solidify Excel's place in the analytics space. Yes, spreadsheets have well known limitations. Spreadsheets are not databases and probably should not be used as one, especially for relational data models. There are better data visualization and exploration tools. There are better dashboarding tools that are more easily maintained in a production environment. Spreadsheet errors can be a \href{https://eusprig.org/research-info/horror-stories/}{huge problem}. Version control is difficult as both data and logic live in the same spreadsheet document. Documentation of multi-step manual data manipulation processes is often lacking -- hindering reproducibility and automation. There are data size limitations and external data links present their \href{https://www.thespreadsheetguru.com/find-remove-external-links/}{own set of challenges}. But, Excel continues to be used for a wide range of quantitative analysis related activities throughout the business world. One particular use is the building and exercising of models of business problems
to explore the impact of important model inputs on different output metrics. This ``what if?'' type modeling is facilitated by Excel's flexibility, numerous built in functions, integrated sensitivity analysis via \href{https://support.microsoft.com/en-us/office/calculate-multiple-results-by-using-a-data-table-e95e2487-6ca6-4413-ad12-77542a5ea50b}{Data Tables},  \href{https://support.microsoft.com/en-us/office/use-goal-seek-to-find-the-result-you-want-by-adjusting-an-input-value-320cb99e-f4a4-417f-b1c3-4f369d6e66c7}{Goal
	Seek} for break-even analysis and exploratory data analysis via plots and Pivot Tables.

@@

Most AAP students have also already taken   \href{http://www.sba.oakland.edu/faculty/isken/courses/ba}{Business Analytics} (BA) which is a spreadsheet based introduction to business analytics \citep{isken2003:muddy,isken2014:lab}. The BA and PCDA courses have historically been offered in face to face mode in a computer teaching lab. Since 2020, these two courses have been also offered as online, asynchronous courses. The newer AAP course has been offered since 2021 in an online asynchronous mode. All three courses have extensive course websites that are publicly accessible, including access to all of the video content and supporting files. These courses serve upper level undergraduate and graduate students. Many are from the School of Business Administration but they draw students from many programs across our campus.

@@

All of the Python modules in the PCDA course as well as the entire AAP course are taught using a combination of Jupyter notebooks, Python scripts, and videos that walk readers through the notebooks. Jupyter notebooks, part of \href{https://jupyter.org/}{Project Jupyter}, are an example of \textit{literate computing} \citep{perez2015project} and allow one to create rich interactive documents that support iterative exploration and development.

Central to Jupyter notebooks is a mix of \textit{code cells}, \textit{markdown cells} and \textit{output cells}. Code cells contain executable code while markdown cells contain plain text using \href{URL}{markdown} for styling text. For example, Figure \ref{fig:cells_edit_mode} shows two markdown cells separated by a code cell as they look when in editing mode. After running the cells, they appear as shown in Figure \ref{fig:cells_after_running}. After executing a code cell, the output appears directly below it in an output cell. Mixing code and markdown cells makes it quite easy to create interactive tutorials in which students can not only run pre-written code, but can also add their own code -- all with the aid of supporting explanatory text. All of the blog posts upon which EwP is based were written in Jupyter notebooks.

The ability to weave standard explanatory language with formal programming languages has made notebook computing a popular tool in the data science education community \citep{perkel2018jupyter}. Jupyter notebooks can also include equations written in \LaTeX, data visualizations, and other multi-media. The technology is free, open source, runs on most computing platforms, and has a huge user community that has created a large number of resources for students, faculty and industry practitioners. There are countless introductory tutorials for Jupyter notebooks, but the official Project Jupyter Documentation \citep{jupyterteamProjectJupyterDocumentation2015} is a very good place to start if you are new to notebook computing. As with all tools, there are pros and cons to their use. See \citet{rule2018ten, barba2019teaching} and \citet{johnson2020benefits} for guidance on Jupyter notebook best practices for both research and teaching.

@@

There are two overarching learning objectives for this \href{http://www.sba.oakland.edu/faculty/isken/courses/aap/mod3a_whatif.html}{first EwP submodule} - see Figure \ref{fig:objectives_whatif} for how the objectives are presented to students via the course website. Students learn how Python might be used to do something they are quite familiar with doing in a spreadsheet -- model building and sensitivity analysis. At the same time, they encounter more advanced Python programming concepts and techniques. In this way, the computational problem to be solved drives the introduction of more advanced Python concepts and techniques instead of such things being presented in a vacuum. I make it a point to show that there are often multiple possible approaches to a computational problem by being willing to abandon one approach and trying another. Emphasis is given to the lessons learned from each path explored. Every notebook is accompanied by screencasts that walk the students through the material and act as a guided tour with challenges to overcome along the way. While all of the notebooks are publicly available (see Section \ref{sec:software}), links to html versions of each of the notebooks are provided throughout the document and readers are encouraged to use them while reading to see exactly how each topic is presented.

@@

The stage is set with an overview of a typical spreadsheet based model that most of the students have seen before. The notebook also includes a screenshot (Figure \ref{fig:xl_model}) of this Excel model.

@@

We then introduce a similar modeling problem that will be the focus of the rest of the notebook. The Walton Bookstore problem appears in the chapter on Monte-Carlo simulation in textbooks that I have used in my spreadsheet modeling course \citep{albrightBusinessAnalyticsData2016,winstonPracticalManagementScience2018}. The (slightly modified) problem is as follows:

@@

At this point we discuss a fundamental difference between the computing model of Excel and that of Jupyter notebooks. While spreadsheets respond automatically to changes, Jupyter notebooks require rerunning all code cells that include or follow the changed code. This difference in behavior between Jupyter notebooks and spreadsheets might be one of the most important concepts for students to internalize. A well known challenge, often referred to as the \textit{hidden state problem}, associated with computational notebooks is the ease of running code cells out of order and introducing inadvertent errors -- see \citet{johnson2020benefits,grusDonNotebooks2018}. This point along with strategies to avoid it are discussed during the first week of the course which focuses on \href{http://www.sba.oakland.edu/faculty/isken/courses/aap/jupyter_notebooks.html}{notebook computing best practices as well as a historical look} at computational notebooks.

@@

Since order quantity is the key decision variable, we might want to see how profit changes for different order quantities. In Excel, the Data Table tool provides a familiar and easy way to do this. In a Data Table, a range of order quantities can be used as row or column input and one or more output values can be computed. In Python, the vectorized nature of the \href{https://numpy.org/}{NumPy} library \citep{harris2020array} provides a simple way to accomplish the same thing. A range of order quantities is created as a NumPy array that can be used directly to compute vectors of all of the intermediate and final output variables. Much like Data Tables, vectorized computations avoid having to explicitly iterate, or loop, through a collection of input values.

@@

Before trying to build our \code{goal\_seek} function we take a brief detour into basic root finding algorithms through packages like \href{https://docs.scipy.org/doc/scipy/tutorial/optimize.html#root-finding}{SciPy} \citep{2020SciPy-NMeth} as well as in \href{https://github.com/patrickwalls/mathematical-python/}{open source notebooks and scripts} that are widely available in the Python data science ecosystem \citep{wallsMathematicalPython2023}. We compare the output of Excel's Goal Seek with different starting values to the values obtained by different root finding algorithms implemented in Python. The importance of starting values and also the range of different root finding algorithms available is often surprising to business students who have spent most of their analytical life in Excel.

@@

In the Excel modeling world, Monte-Carlo simulation can be done without add-ins, but packages like \href{https://lumivero.com/products/at-risk/}{@Risk} can make the process much easier. In particular, random variable generation for a wide range of probability distributions is facilitated by add-ins like @Risk. In Python, the \href{https://numpy.org/doc/stable/reference/random/generator.html}{NumPy package provide random variate generation} for a large number of distributions and similarly, \href{https://docs.scipy.org/doc/scipy/reference/stats.html}{SciPy provides functions for computing distribution related quantities} for many probability distributions. Much like we did when creating the \code{data\_table} function, we start with adding uncertainty to a single input variable in the bookstore model and rely on NumPy's vectorized computing capabilities to generate the profit associated with each realization of the random input. Then we can do standard statistical analysis of the simulation output using Python packages such as pandas and SciPy. In spreadsheet based simulation textbooks, the notion of the ``Flaw of Averages'' \citep{savageFlawAveragesWhy2012} is often illustrated. We do the same with our Python based model, showing how replacing random demand by its mean can result in a wildly optimistic estimate of the mean profit found through analyzing the simulation output. Much like doing simulation in Excel without add-ins, using Python makes transparent the quite simple and brute force nature of Monte-Carlo simulation.

@@

Now that we have created some useful modeling functionality, our goal is to make it easier for a modeler to actually use the code. Certainly we do not want to rely on copying, pasting and modifying code when building a new model. In the Python world, the natural next step is to deploy our code as a Python \textit{package} \citep{buezenWelcomePythonPackages2022} that can then be imported like any other Python library. Figure \ref{fig:objectives_packaging} shows the learning objectives from this \href{http://www.sba.oakland.edu/faculty/isken/courses/aap/mod3b_whatif_packaging.html}{submodule page on the course website}.

@@

Creating a deployable Python package is a good exercise in designing and creating a reusable software artifact. Thinking about how others will interact with your software usually leads to better designed software.  In this submodule we create a Python package that can be imported and used by other modelers. This will require some design changes to our software and will also involve moving code out of Jupyter notebooks and into Python script files. While Jupyter notebooks can be integrated into production environments \citep{uffordInteractiveNotebookInnovation2019}, it is much more common for code to live in Python script files. In addition to creating a deployable package, this submodule also discusses the importance of documentation in its many forms in a Python project.

@@

The first notebook in this submodule starts by explaining what Python packages are and their role in sharing code. We then revisit a concept from earlier in the course -- creating a good project folder structure. For this we use what is known as a \href{https://cookiecutter.readthedocs.io/en/stable/}{cookiecutter} that can automatically generate a project folder structure and key files from the answers to a few prompts \citep{audreyroyandcookiecuttercommunityCookiecutterBetterProject2012}. There are a few different folder layouts used in Python projects and there is a bit of discussion on the folder structure we will use and resources given for exploring this issue further. Students are reminded of the importance of version control and we initialize a git repository for our budding package. The basics of version control were already covered in the first week of the course. My experience is that most students have little exposure to software project management concepts such as project folder structures and version control. This can lead to a jumble of files with incoherent names used as a proxy for version control -- \code{whatif\_v1.py}, \code{whatif\_v2.py}, \code{whatif\_final.py}, \code{whatif\_finalfinal.py}, and so on.

@@

This notebook ends by pointing the students to another notebook in which our newly deployed whatif package is used for a completely different model -- the New Car Simulation that has been a part of Winston and Albright's textbooks for many years \citep{winstonPracticalManagementScience2018,albrightBusinessAnalyticsData2016}. This new model differs structurally from the Bookstore Model, but is representative of many financial models in that cash flows over multiple years need to be computed and summarized. This helps cement the idea that there is value in creating reusable code.

@@

The deployment focused submodule ends with a short notebook discussing the different types of documentation needed in a typical project. This includes code comments, \href{https://peps.python.org/pep-0257/}{docstrings}, \code{readme} files and generating documentation by creating \href{https://www.sphinx-doc.org/en/master/usage/restructuredtext/basics.html}{reStructuredText} files (similar to but more powerful than markdown) and using \href{https://www.sphinx-doc.org/en/master/index.html}{Sphinx}, the Python documentation generation tool. This is one of those tasks that few people like to do but can be very important for the long term success of a project. Even if you are just creating a tool for yourself, you will forget the details over time and documentation can be very beneficial in refreshing your memory.
